\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-12 paper.tex
Unary nodes

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do much more
besides removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
revolves around the presence of nodes that appear
``unary'' in one or more local trees
(where a node has exactly one child over a particular genomic region).
These arise in several different ways.
Nodes with one child are not a standard feature of evolutionary trees;
we are usually interested in
nodes that have at least two children, because these
are, in principle, detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals
% Following feedback from Wilder, this is more likely to confuse
% people unless we properly explain
% (or indeed cell divisions)
and there is usually
very little information about these intermediates, or utility
in modelling them.
% FIXME the bridge between the next few paragraphs is a bit weird, in
% that we've just been talking about how interesting unary nodes are
% but then go on about how to remove them. Just needs a bit of smoothing
% out I think.
ARGs, however, are different, and there
can be significant information about evolutionary processes
provided by nodes that are unary in one or more local trees.
% Not sure what to do with this - should mention though.
\citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.''
% Full quote from fig caption:
% Genetic ancestry in the form of the ARG for a single individual. Combining
% genetic ancestry from different positions leads to a graph, incorporating all
% realized genetic ancestry paths, implicitly passing through points representing
% specific individuals.

@@

Fig.~\ref{fig-simplification} shows a series of successive steps of
simplification,
% FIXME this doesn't explain anything - what did we store in the simulation?
starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.
In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify,
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.

@@

% JK: Text moved from an earlier section. See if we want to reuse this here
% if not already said.
% Indeed a single node can have both multiple children
% and multiple parents. This would be the case in Fig~\ref{fig-arg-in-pedigree} if,
% for instance, node \noderef{a} were to have two children.
% % It is a shame that this is not actually shown in fig 3, but I don't think
% % we can do this unless we have 5 generations
% Moreover, a single node can have more than 2 children, and (as we shall see in
% the ``\nameref{ARG_simplification}'' section), more than 2 parents. Another way to
% state this is that transitions between nodes in a gARG can encapsulate multiple
% events in an eARG.

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all other local trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees. The extent to which this information can be inferred
and utilised is an interesting open question.
% I don't think there's an actual manuscript or title to refer to here
% [CITE Ralph in prep.].

>>>>

Nodes with one child are not a standard feature of evolutionary trees
or the ARG literature
(although \citet{mathieson2020ancestry} show examples in which the ARGs
contain unary nodes, and state that paths are
``implicitly passing through points representing specific individuals.'').
We are usually interested in
nodes that have at least two children because these
are in principle detectable from mutational differences
arising in the respective subtrees.
A tree branch may correspond to many generations
of individuals (or indeed cell divisions; see Appendix XXX)
and there is usually
little information about these intermediates, or utility
in modelling them.
As we see in the next section, however, there is a range of
different ways in which nodes can be unary in the local trees
of an ARG. Furthermore, ARGs inferred by current methods contain
locally unary nodes from several of them (see section XXX).

\section*{Levels of simplification}\label{ARG_simplification}
ARG simplification is a powerful tool, and can do far more
than removing unreachable nodes and edges in a prospective ARG
and resolving ancestral material in an eARG.
In general, we can think of
ARG simplification as the process
of removing nodes and re-writing edges (and their inheritance annotations)
to remove various types of redundancy.
The redundancy that we are interested in
% We are interested in both graph and local unary, so qualifying this
% would just make it less readable, IMO
revolves around the presence unary nodes (see previous section),
which we illustrate by an example simulation.
[THIS SIMULATION...]

@@

Fig.~\ref{fig-simplification} shows a series of successive steps of
simplification,
% FIXME this doesn't explain anything - what did we store in the simulation?
starting with a sample-resolved but unsimplified gARG, simulated under a
backward-time Wright-Fisher
model. We show the graph on the left, and the corresponding sequence of
local trees on the right.

In Fig.~\ref{fig-simplification}A the local trees (right)
contain many unary nodes, and as we successively simplify,
the local trees contain fewer
unary nodes (Fig.~\ref{fig-simplification}B,C) until
we reach Fig.~\ref{fig-simplification}D, where the local trees
have no unary nodes.
A node is unary in a local tree covering the genome interval $I$
if genetic inheritance passes through that ARG node,
but no coalescence occurs in the interval $I$.
The distinction between the ``common ancestry'' of two or more genomes
in an ancestral genome and the ``coalescence'' which may or may
not occur in the local trees is
important~\citep{hudson1983testing,kelleher2016efficient}.
Consider \noderef{e} in Fig.~\ref{fig-simplification}A,
for example. We can see from the graph that it is a common
ancestor of  samples \noderef{a} and \noderef{b}, but
it does not correspond to any coalescence in the
local trees to the left of position $44$, and is therefore
unary in these three trees.

@@

The remaining nodes are MRCAs of some subset of the samples
at \emph{some} positions along the genome. We still have
some unary nodes in the local trees, but these nodes will
correspond to a coalescence in at least one other
local tree. For example, node  \noderef{k} is unary in the fourth tree
of Fig.~\ref{fig-simplification}C, but is either binary
or ternary in all other local trees (recall this is a Wright-Fisher
simulation). The final level of simplification is to alter the edge annotations
such that, although no nodes are removed from the graph, all
unary nodes disappear from the local trees (Fig.~\ref{fig-simplification}D).
Note that although this last stage produces simpler local trees, by
removing information about the exact paths taken by lineages through
the graph, we lose potentially useful information about shared edges
between trees.
The \texttt{msprime} simulator, and the version of Hudson's algorithm described
by~\citet{kelleher2016efficient}, produces ARGs
that are fully simplified (i.e., contain no locally unary nodes).
It is not difficult, however, to update
these methods to record information about the passage of ancestral
material through genomes under a range conditions.
